---
title: "DEBATE WORDS"
author: "Winston Saunders"
date: "November 18, 2015"
output: 
    html_document:
        theme: united
---

###SUMMARY
This analysis is based on the premise that widening political divisions are not just about "what to do" but, more fundamentally, about "what the actual problem is." For instance even the conservative [National Review](http://www.brookings.edu/blogs/fixgov/posts/2015/06/03-iowa-bloomberg-galston) highlighted disagreement about the most important issues.  
Here, using standard NLP (Natural Lanuguage Processing) techniques, I explore this question looking for differences in the lanuguage of presidential candidates using texts from recent Republican and Democratic debates. 
Key findings are:  
1. "Wordcloud" visualization shows systematic differences between candidates, though more surprising are the similarities.  
2. Word frquency analysis highlights positional differences between candidatates, but does not convey important context information.  
3. Bigram toeknization and word-stem searches do the best job of revealing position differences between candidates.  


###DATA SOURCES AND METHODS
The text of the presidential debates are downloaded from the [UCSB Presidency Project](http://www.presidency.ucsb.edu/debates.php). Transcripts were pasted into Apple Pages and stored as unformated .txt files. From that point all processing is done with __R__ using capabilities of {tm} and associated libraries.  

  

###CANDIDATE WORD-CLOUDS

The quickest and most visual method to compare texts is word-frequency analysis using wordclouds. 
Not surprisingly, word choices vary significantly between candidates. However, there are also some striking similarities.  

```{r, echo=FALSE, warning=FALSE, message=FALSE}


load_debate_text <- function(file_name){

## GET THE DATA
    
    ## data are the raw text files downloaded from the UCSB website.
    
    directory <- "/Users/winstonsaunders/Documents/Presidential_Debates_2015/"
    mydata <- read.table(paste0(directory, file_name), header=FALSE, sep="\n", stringsAsFactors = FALSE, quote = "")

    ## This is what a sample of the the raw data looks like
    ## mydata[100:105,1]
        # [1] "QUINTANILLA: Hold it. We'll cut it back to..."
        # [2] "QUICK: Dr. Carson, let's talk about taxes."
        # [3] "You have a flat tax plan of 10 percent flat taxes..."
        # [4] "If you were to took a 10 percent tax..."
        # [5] "So what analysis got you to the point where you think this will work?"
        # [6] "CARSON: Well, first of all..."

## ADD COLUMNS OF NUMBERS AND SEPARATE TEXT FROM CANDIDATE NAMES
    ## Add number column
    #mydata$n<-1:nrow(mydata)
    
    ## get rid of "..."
    #mydata$V1<-gsub("...", "", mydata$V1)
    
    ## This regex gets rid of all but the capitalized names
    mydata$name <- gsub(":.+|^[A-Z][^A-Z].+", "", mydata$V1 )

    ## Fill in the blank rows
    for (i in 2: nrow(mydata)){
        if (mydata$name[i] == "") mydata$name[i] <- mydata$name[i-1]
    }

    ## CREATE COLUMN OF DEBATE TEXT AND CLEAN UP BLANK SPACES
    mydata$text <- gsub (".", "", mydata$V1, fixed=TRUE)
    mydata$text <- gsub ("'", "", mydata$text)
    mydata$text <- gsub ("=", "", mydata$text)
    mydata$text <- gsub ("[A-Z]{2,}: ", "", mydata$text)
    mydata$text <- tolower(mydata$text)
    mydata$text <- gsub ("   ", " ", mydata$text)
    mydata$text <- gsub ("  ", " ", mydata$text)
    ##stem a few words
    mydata$text <- gsub ("taxes", "tax", mydata$text)
    mydata$text <- gsub ("guns", "gun", mydata$text)
    mydata$text <- gsub ("veterans", "veteran", mydata$text)
    mydata$text <- gsub ("terrorists", "terror", mydata$text)
    mydata$text <- gsub ("terrorism", "terror", mydata$text)
    
    #mydata$text <- gsub ("americans", "american", mydata$text)
    #mydata$text <- gsub ("american", "america", mydata$text)

    
    
    
    ## the data frame now contains four columns which look like this...
    
        #   50 Our greatest days lie ahead...
        #   51 QUINTANILLA: Mr. Trump?
        #   52 TRUMP: I think maybe ...[laughter]
        #     n        name
        # 50 50       RUBIO
        # 51 51 QUINTANILLA
        # 52 52       TRUMP
        # text
        # 50  Our greatest days lie ahead ....
        # 51  Mr. Trump?
        # 52  I think maybe my greatest weakness ... [laughter]
    
    return(mydata)
}

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

candidate_text <- function(candidate, mydata){
    ## 
    ## GET CANDIDATE TEXT
    ##
    ## Assumes load_debate_text has been run and the debate text is stored in "mydata"
    ## creates a text list
    
  
    
    ## filter for candidate
    text<-mydata$text[mydata$name==candidate]
    text<-paste(text, collapse = " ")

    return(text)


}
```


```{r echo=13:22, warning=FALSE, message=FALSE}

    library(tm)
    library(RWeka)

text_tc <- function(mydata){
    ## 
    ## 
    ##
    ## Assumes text is stored in "mydata"
    ## creates a Corpus from the text
    ## filters some words out 

    require(tm)
    require(SnowballC)

    t_c <- Corpus(VectorSource(text))
    t_c <- tm_map(t_c, content_transformer(tolower))
    t_c <- tm_map(t_c, removePunctuation)
    t_c <- tm_map(t_c, removeNumbers)
    t_c <- tm_map(t_c, removeWords, stopwords("english"))
    t_c <- tm_map(t_c, removeWords, c("applause", "thats", "laughter", "dont", "back", "can", "get", "cant", "come", "big", "inaudible"))

    return(t_c)   
    
} 

```


```{r echo=FALSE, warning=FALSE, message=FALSE}

    library(tm)
    library(RWeka)
    library(SnowballC)

candidate_text_tc <- function(candidate, mydata){
    ## 
    ## GET CANDIDATE DATA
    ##
    ## Assumes load_debate_text has been run and the debate text is stored in "mydata"
    ## creates a Corpus from the candidate text
    
  
    
    ## filter for candidate
    text<-mydata$text[mydata$name==candidate]
    text<-paste(text, collapse = " ")

    require(tm)
    require(SnowballC)

    t_c <- Corpus(VectorSource(text))
    t_c <- tm_map(t_c, content_transformer(tolower))
    t_c <- tm_map(t_c, removePunctuation)
    t_c <- tm_map(t_c, removeNumbers)
    t_c <- tm_map(t_c, removeWords, stopwords("english"))
    t_c <- tm_map(t_c, removeWords, c("applause", "thats", "laughter", "dont", "back", "can", "get", "cant", "come", "big", "inaudible"))

    #t_c <- tm_map(t_c, stemDocument)
    
    return(t_c)   
    
} 

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

library(wordcloud)
library(RColorBrewer)

 color_map2<-c("#00003B","#041851","#4E0812" ,"#051E65", "#650A16", "#103374",  "#9F1A2D", "#482B25")

c_wordcloud <- function(t_c){
    
    ## 
    ## CREATE WORD CLOUD
    ##
    ## Assumes a text Corpus has been created
    ## 
    
    set.seed(8675309)
    
   
    color_map2<-c("#00003B","#041851","#4E0812" ,"#051E65", "#650A16", "#103374",  "#9F1A2D", "#482B25")
    
    wordcloud(t_c, scale=c(4,0.4), max.words=10000, min.freq=5, random.order=FALSE, rot.per=0.2, use.r.layout=FALSE, colors=color_map2)

    

}

```



```{r, echo=FALSE, warning=FALSE, message=FALSE}

    ## GET ALL REPUB DEBATES
     file_name<-"Republican Candidates Debate in Boulder Colorado October 28 2015.txt"
     r_oct<-load_debate_text(file_name)
     #r_oct$date<-"Oct"
     file_name<-"Republican Candidates Debate in Milwaukee Wisconsin November 10 2015.txt"
     r_nov<-load_debate_text(file_name)
     #r_nov$date<-"Nov"
    
    r_all = rbind(r_oct, r_nov)
        ## slim down data a bit
        ## reduce data to about 30%
#         r_all<-r_all[grepl("RUBIO", r_all$name)|
#                          grepl("CARSON", r_all$name)|
#                          grepl("TRUMP", r_all$name)|
#                          grepl("CRUZ", r_all$name)|
#                          grepl("HUCKABEE", r_all$name)|
#                          grepl("FIORINA", r_all$name), ]
    
    
    
    ## GET ALL DEMO DEBATES
     file_name<-"Democratic Candidates Debate in Las Vegas Nevada October 13 2015 .txt"
     d_oct<-load_debate_text(file_name)
     file_name<-"Democratic Candidates Debate in Des Monies Iowa November 14 2015 .txt"
     d_nov<-load_debate_text(file_name)
    
    d_all = rbind(d_oct, d_nov)
    
    ## CREATE TCs FOR EACH CANDIDATE
    trump_all<-candidate_text_tc("TRUMP",r_all)
    rubio_all<-candidate_text_tc("RUBIO",r_all)
    fiorina_all<-candidate_text_tc("FIORINA",r_all)
    carson_all<-candidate_text_tc("CARSON",r_all)
    cruz_all<-candidate_text_tc("CRUZ",r_all)
    huckabee_all<-candidate_text_tc("HUCKABEE",r_all)
    bush_all<-candidate_text_tc("BUSH",r_all)

    clinton_all<-candidate_text_tc("CLINTON",d_all)
    sanders_all<-candidate_text_tc("SANDERS",d_all)
    
    
    
```


<style>
  .col2 {
    columns: 2 300px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 300px; /* chrome, safari */
    -moz-columns: 2 300px;    /* firefox */
  }
  .col3 {
    columns: 3 200px;
    -webkit-columns: 3 200px;
    -moz-columns: 3 200px;
  }
</style>




Let's first just look at the word clouds of different candidates. We can address whether there are differences in the word frequencies used by candidates, as emphasized by the algorithm of the {wordcloud} package. Are there differences within the same party, between candidates, etc.? 

####THE POPULISTS: TRUMP AND SANDERS

Here are the word cloud of Donald Trump's and Bernie Sanders's dialogue at the debates. It's surprising that their frequent word choices like _people_, _country_, and _going_ are common, as if they are both painting a vision of the future that is personal (though radically different in nature).

<div class="col2">


```{r, echo=TRUE, message=FALSE, fig.align="center", results="asis"}
c_wordcloud(trump_all)
c_wordcloud(sanders_all)
```

</div>

####THE WOMEN: HILARY AND CARLY

In this case word clouds couldn't be more different. Hilary's emphasizes _think_ and _people_ while Carly's, a former business woman, primarily emphasizes _government_. 

<div class="col2">


```{r, echo=TRUE, message=FALSE, fig.align="center", results="asis"}
c_wordcloud(clinton_all)
c_wordcloud(fiorina_all)
```

</div>

####BRAINS VERSUS BRAUN: CRUZ AND HUCKABEE

Ted Cruz's wordcloud emphasizes wonkish technicalities, like _taxes_ and _washington_, while that of Mike Huckabee, a former minister, mixes the populist and a focus on  _government_. 

<div class="col2">


```{r, echo=TRUE, message=FALSE, fig.align="center", results="asis"}
c_wordcloud(cruz_all)
c_wordcloud(huckabee_all)
```

</div>


###STAYING ON MESSAGE: COMPARING DEBATES

We can also split the text by debate. Since the debates cover different topics and questions, one might expect to see this reflected in the text of the separate dialogues. What's surprising here is how comparable the language of each candidate is between the debates. Perhaps the candidates are more interested in staying on message than answering questions directly?

<div class="col2">


```{r, echo=TRUE, message=FALSE, fig.align="center", results="asis"}
c_wordcloud(candidate_text_tc("TRUMP", r_oct))
c_wordcloud(candidate_text_tc("TRUMP", r_nov))
```
</div>

<div class="col2">


```{r, echo=TRUE, message=FALSE, fig.align="center", results="asis"}
c_wordcloud(candidate_text_tc("SANDERS", d_oct))
c_wordcloud(candidate_text_tc("SANDERS", d_nov))
```
</div>

__A NOTE ON WORDCLOUDS__ Both visually appealing and interesting, wordclouds reveal differences between candidate word choices which hint at political preferences and opinions, but do not reveal significant detail about them. For instance, differences in positions on policy and important matters like taxes, terrorism, immigration, and class division are not deducible from the wordclouds.  

###WORD FREQUENCY

We can check word frequency directly by simply tokenizing the text and counting single words. Looking for the most frequent words used by each candidate may reveal some clearer differences. To do this analysis some additional words like "thats", "dont", "back", "can", "get", "cant", and "come" are suppressed.  


<style>

table { 
    width: 80%;
    display: table;
    border-collapse: collapse;
    border-spacing: 18px;
    border-color: #AAAAFF;
    background-color: #D5CAB6;
    text-align: center;
    padding: 2px;
    font: 16px arial, sans-serif;
}
th{
    text-align: center;
}, td {
    text-align: left;
    padding: 6px;
}
</style>


```{r, echo=FALSE}

## Create Term_Document_Matrices

TDM_trump <- TermDocumentMatrix(trump_all)
TDM_rubio <- TermDocumentMatrix(rubio_all)
TDM_fiorina <- TermDocumentMatrix(fiorina_all)
TDM_carson <- TermDocumentMatrix(carson_all)
TDM_cruz <- TermDocumentMatrix(cruz_all)
TDM_huckabee <- TermDocumentMatrix(huckabee_all)
TDM_bush <- TermDocumentMatrix(bush_all)

TDM_clinton <- TermDocumentMatrix(clinton_all)
TDM_sanders <- TermDocumentMatrix(sanders_all)

```



```{r, echo=FALSE, results='asis', fig.align='center'}

library(xtable)

## Create table of frequent terms
## note: The Frequent Term utility does not rank by order. It produces a character vector of terms in x which occur more or equal often than lowfreq times and less or equal often than highfreq times.

# row_names <- c("Trump", "Sanders", "Clinton", "Fiorina", "Cruz", "Rubio" )
# col_names <- c("most frequent", "Second", "Third", "Fourth", "Fifth")

# word_mat<-matrix(
#     c(findFreqTerms(TDM_trump, 30)[1:5],
#       findFreqTerms(TDM_sanders, 30)[1:5],
#       findFreqTerms(TDM_clinton, 30)[1:5],
#       findFreqTerms(TDM_fiorina, 10)[1:5],
#       findFreqTerms(TDM_cruz, 10)[1:5],
#       findFreqTerms(TDM_rubio, 10)[1:5]), nrow = 6, byrow = TRUE, dimnames = list(row_names ,col_names ) )
# 
# word_df<-as.data.frame(word_mat)
# 
# 
# 
# print(xtable(word_df), type='html', comment=FALSE, include.rownames=TRUE, 
#       html.table.attributes='border="3" align="center" ' )


```


These tell a bit of a clearer story, i.e. we can almost read these words like sentences or sentence fragments. Both Bernie Sanders and Donald Trump again seems to come across similiarly as populists. The first ranking word for Donald Trump is _country_ and for Bernie Sanders it's _believe_. It's interesting to note that the notion of getting the "country going" comes through in the top three candidates.  
More humorously, note that the word "clinton" figures in both Carly Fiorina's and Hilary Clinton's lists. These texts almost read as an assertion by Hilary and an counter argument by Carly.  
Ted Cruz's frequent words again seem to focus on business and Carlo Rubio loves America.  
  
While word frequencies reveal differences between the approaches and personalities of the candidates, they don't by themsleves elucidate differences on specific policies or attitudes. Let's try something else.

```{r, echo=FALSE}

#findAssocs_candidate(TDM_trump,"country")
# findAssocs_candidate(TDM_sanders,"people")
# findAssocs_candidate(TDM_clinton,"people")
# findAssocs_candidate(TDM_fiorina,"people")

```




```{r, echo=FALSE}
## Build a matrix owith counts of specific words for a list of candidates

## First convert TDMs to Data Frames

a<-as.matrix(TDM_trump)
b<-as.data.frame(a)
df_trump<-b
colnames(df_trump)<-"trump"

a<-as.matrix(TDM_sanders)
b<-as.data.frame(a)
df_sanders<-b
colnames(df_sanders)<-"sanders"

a<-as.matrix(TDM_fiorina)
b<-as.data.frame(a)
df_fiorina<-b
colnames(df_fiorina)<-"fiorina"

a<-as.matrix(TDM_clinton)
b<-as.data.frame(a)
df_clinton<-b
colnames(df_clinton)<-"clinton"

## merge the data frames

merged_candidates<-merge(df_trump, df_sanders, by=0, all=TRUE)
rownames(merged_candidates) <- merged_candidates$Row.names
merged_candidates$Row.names <- NULL
merged_candidates<-merge(merged_candidates, df_clinton, by="row.names", all=TRUE)
rownames(merged_candidates)<-merged_candidates$Row.names
merged_candidates$Row.names <- NULL
merged_candidates<-merge(merged_candidates, df_fiorina, by="row.names", all=TRUE)


## fix NAs
merged_candidates[is.na(merged_candidates)]<-0

##Get rid of self references
merged_candidates["clinton", "clinton"]<-0
merged_candidates["trump", "trump"]<-0
merged_candidates["sanders", "sanders"]<-0
merged_candidates["fiorina", "fiorina"]<-0

## sum all
merged_candidates$all <- merged_candidates$trump + merged_candidates$clinton + merged_candidates$sanders + merged_candidates$fiorina

## sort it
merged_candidates<-merged_candidates[with(merged_candidates, order(-all)), ]

#merged_candidates <- merged_candidates[merged_candidates$all>50,]
## convert Row.names to a factor
merged_candidates$Row.names<-as.factor(merged_candidates$Row.names)

merged_candidates<-merged_candidates[complete.cases(merged_candidates),]

head(merged_candidates)


## This is what the data look like
#      Row.names trump sanders clinton fiorina all
# 1810    people    33      85      53      10 181
# 2483     think     9      55      90       9 163
# 1050     going    44      44      45      10 143
# 561    country    34      70      25       1 130
# 1363      know    23      26      56      19 124
# 2704      well     9      31      56       8 104

```

```{r, echo=FALSE, results="asis"}

library(xtable)

c_1<-merged_candidates[with(merged_candidates, order(-clinton)), ]
s_1<-merged_candidates[with(merged_candidates, order(-sanders)), ]
f_1<-merged_candidates[with(merged_candidates, order(-fiorina)), ]
t_1<-merged_candidates[with(merged_candidates, order(-trump)), ]

compare_mf <- rbind(c_1[1:4,], f_1[1:4,], s_1[1:4,], t_1[1:4,])
colnames(compare_mf)<-c("word", "clinton", "fionrina", "sanders", "trump")


print(xtable(unique(compare_mf), digits=0), type='html', comment=FALSE, include.rownames=FALSE, 
      html.table.attributes='border="3", align="center" ' )

```

####GRAPHICAL REPRESENTATION

There's additional information in whether words used frequently by one candidate are used at all by another candidate. While the wordlcloud analysis gives some insight, we can analysis the data graphically for more quantitative information. Here is a graph of the "top" words used by all candidates

```{r, echo=FALSE, fig.align="center", fig.height= 5, fig.width=8, message=FALSE, warning=FALSE}

library(reshape2)

merged_plot<-melt(merged_candidates[merged_candidates$all>50,])
colnames(merged_plot)<-c("words", "candidate", "count")

library(ggplot2)

# p <- ggplot(merged_plot, aes(x = words, y = count, color=candidate))
# p <- p + geom_point(size=3, pch=19)
# p <- p + theme(axis.text.x=element_text(face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"))
# p <- p + coord_flip()
# p <- p + ggtitle("CANDIDATE WORD COUNT")
# p <- p + xlab("word")
# p <- p + ylab("count")
# p 

```
```{r, echo=FALSE, fig.align="center", fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
mc<-merged_candidates
mc$rank <- 1:nrow(mc)
##normalize
mc$clinton<-mc$clinton/sum(mc$clinton)
mc$sanders<-mc$sanders/sum(mc$sanders)
mc$trump<-mc$trump/sum(mc$trump)
mc$fiorina<-mc$fiorina/sum(mc$fiorina)

mc_plot<-mc[mc$all>40,]

p <- ggplot(mc_plot, aes(x = rank, size=3))
#p <- p + geom_line(size=0.5)
p <- p + geom_point(aes(y = clinton, color="clinton"), size=3)
p <- p + geom_point(aes(y = sanders, color="sanders"), size=3)
p <- p + geom_point(aes(y = trump, color="trump"), size=3)
p <- p + geom_point(aes(y = fiorina, color="fiorina"), size=3)
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1, face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"),legend.title=element_blank())
#p <- p + coord_flip()
p <- p + scale_x_discrete(breaks=1:length(mc$Row.names),
        labels=mc$Row.names)
p <- p + ggtitle("MOST FREQUENT WORD COUNT")
p <- p + xlab("")
p <- p + ylab("normalized frequency")
p 

```

####NORMALIZED Z STATISTICS

```{r, echo=FALSE, fig.align="center", fig.height=4, fig.width=8, message=FALSE, warning=FALSE}

##calculate mean

mc_plot$mean <- rowSums(mc_plot[,c(2,3,4,5)])/4

## subtract
mc_plot$sanders<-(mc_plot$sanders - mc_plot$mean)
mc_plot$clinton<-(mc_plot$clinton - mc_plot$mean)
mc_plot$fiorina<-(mc_plot$fiorina - mc_plot$mean)
mc_plot$trump<-(mc_plot$trump - mc_plot$mean)

## calc squared dev
mc_plot$sdev<- sqrt( ((mc_plot$sanders)^2 + (mc_plot$clinton)^2 + (mc_plot$fiorina)^2 + (mc_plot$trump)^2)/(4-1))

mc_plot$sanders<-(mc_plot$sanders/mc_plot$sdev)
mc_plot$clinton<-(mc_plot$clinton/mc_plot$sdev)
mc_plot$fiorina<-(mc_plot$fiorina/mc_plot$sdev)
mc_plot$trump<-(mc_plot$trump/mc_plot$sdev)



# p <- ggplot(mc_plot, aes(x = rank, size=3))
# #p <- p + geom_line(size=0.5)
# p <- p + geom_point(aes(y = clinton, color="clinton"), size=3)
# p <- p + geom_point(aes(y = sanders, color="sanders"), size=3)
# p <- p + geom_point(aes(y = trump, color="trump"), size=3)
# p <- p + geom_point(aes(y = fiorina, color="fiorina"), size=3)
# p <- p + theme(axis.text.x=element_text(angle=90, hjust=1, face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"),legend.title=element_blank())
# #p <- p + coord_flip()
# p <- p + scale_x_discrete(breaks=1:length(mc$Row.names),
#         labels=mc$Row.names)
# p <- p + ggtitle("Z_STATISTICS MOST FREQUENT WORDS")
# p <- p + xlab("")
# p <- p + ylab("normalized frequency")
# p 

```

The above doesn't reveal much more information than the wordcloud analysis does. However, we can also pick some "key words" and sample for their frequency. For a first stab, let's try


```{r, echo = 6:7, fig.align="center", fig.height=4, fig.width=6, message=FALSE, warning=FALSE }

library(reshape2)

merged_plot<-melt(merged_candidates)
colnames(merged_plot)<-c("words", "candidate", "count")

key_words = c("tax", "government", "climate", "class", "wall", "street","terror", "economy", "immigrant", "america", "veteran", "drug", "health", "gun", "education", "bankruptcy", "money", "women", "war", "rights", "abortion", "violence")
merged_plot<-merged_plot[merged_plot$word %in% key_words, ]
merged_plot<-merged_plot[!merged_plot$candidate=="all",]




p <- ggplot(merged_plot, aes(x = words, y = count))
p <- p + facet_grid(.~candidate)
p <- p + geom_point(size=3, pch=19, color = "#228222")
p <- p + theme(axis.text.x=element_text(face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"))
p <- p + coord_flip()
p <- p + ggtitle("DIFFERENCES IN CANDIDATE WORD CHOICE")
p <- p + xlab("")
p <- p + ylab("word count")
p 



mc<-mc[mc$Row.names %in% key_words,]
mc$rank<-1:nrow(mc)

head(mc)

p <- ggplot(mc, aes(x = rank, size=3))
#p <- p + geom_line(size=0.5)
p <- p + geom_point(aes(y = clinton, color="clinton"), size=3)
p <- p + geom_point(aes(y = sanders, color="sanders"), size=3)
p <- p + geom_point(aes(y = trump, color="trump"), size=3)
p <- p + geom_point(aes(y = fiorina, color="fiorina"), size=3)
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1,face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"),legend.title=element_blank())
#p <- p + coord_flip()
p <- p + scale_x_discrete(breaks=1:25,
        labels=mc$Row.names)
p <- p + ggtitle("CANDIDATE WORD COUNT")
p <- p + xlab("")
p <- p + ylab("normalized frequency")
p 

```

```{r, echo=FALSE, fig.align="center", fig.height=4, fig.width=8, message=FALSE, warning=FALSE}

##calculate mean

mc_plot<-mc

mc_plot$mean <- rowSums(mc_plot[,c(2,3,4,5)])/4

## subtract
mc_plot$sanders<-(mc_plot$sanders - mc_plot$mean)
mc_plot$clinton<-(mc_plot$clinton - mc_plot$mean)
mc_plot$fiorina<-(mc_plot$fiorina - mc_plot$mean)
mc_plot$trump<-(mc_plot$trump - mc_plot$mean)

## calc squared dev
mc_plot$sdev<- sqrt( ((mc_plot$sanders)^2 + (mc_plot$clinton)^2 + (mc_plot$fiorina)^2 + (mc_plot$trump)^2)/(4-1))

mc_plot$sanders<-(mc_plot$sanders/mc_plot$sdev)
mc_plot$clinton<-(mc_plot$clinton/mc_plot$sdev)
mc_plot$fiorina<-(mc_plot$fiorina/mc_plot$sdev)
mc_plot$trump<-(mc_plot$trump/mc_plot$sdev)



p <- ggplot(mc_plot, aes(x = rank, size=3))
#p <- p + geom_line(size=0.5)
p <- p + geom_point(aes(y = clinton, color="clinton"), size=3)
p <- p + geom_point(aes(y = sanders, color="sanders"), size=3)
p <- p + geom_point(aes(y = trump, color="trump"), size=3)
p <- p + geom_point(aes(y = fiorina, color="fiorina"), size=3)
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1, face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"),legend.title=element_blank())
#p <- p + coord_flip()
p <- p + scale_x_discrete(breaks=1:length(mc$Row.names),
        labels=mc$Row.names)
p <- p + ggtitle("Z_STATISTICS MOST FREQUENT WORDS")
p <- p + xlab("")
p <- p + ylab("normalized frequency")
p 

```

###WORD ASSOCIATIONS FROM BIGRAM TOKENIZATION

Since word fequency does not convey specific positions on issues, let's look at word associations to see if we can get closer to meaning from more information about the context of word usage. This analysis simply tokenizes the text as bigrams, then uses a simple function

```{r eval=FALSE}
bigram_table[grep(word, rownames(bigram_table), ignore.case=TRUE)]
```

to pull out relevant terms from the torkenized TDM.
A key challenge is that the texts are relatively short, so statistics comparing the word frequencies are poor. Nevertheless, we can see that context around different words, even at the relatively unsophisticated level of simple bigrams, starts to hint at differences in approach to problems. 

```{r, echo=FALSE}

library(RWeka)
bigramTokenizer <- function(x) NGramTokenizer(x, 
                                Weka_control(min = 2, max = 2))

trump_bigrams <- sort(table(bigramTokenizer(trump_all)), decreasing = TRUE)
sanders_bigrams <- sort(table(bigramTokenizer(sanders_all)), decreasing = TRUE)
clinton_bigrams <- sort(table(bigramTokenizer(clinton_all)), decreasing = TRUE)
fiorina_bigrams <- sort(table(bigramTokenizer(fiorina_all)), decreasing = TRUE)

bigram_finder <- function(word, bigram_table){
    relevant_bigrams <- bigram_table[grep(word, rownames(bigram_table), ignore.case=TRUE)]
    return(relevant_bigrams)
}

```



###CANDIDATES ON TAXES

Bernie talks about _"tax"_ and _"terror"_ as well. His discussion of taxes has a reformist bent, but where Carly Fiorina talks associates words like budgeting, changes, reform, simplify, code, reform, and plan, Bernie Sanders associates words like cap, income, must, share, speculation, breaks, reform, wall, and rebuilding. 




```{r, echo=FALSE, fig.align="center", fig.height=4, fig.width=6, warning=FALSE, message=FALSE}


word_sample <- "tax"
a<-bigram_finder(word_sample, fiorina_bigrams)
a<-as.data.frame(a)
colnames(a)<-word_sample

bigram_freq<-a

library(ggplot2)

p <- ggplot(bigram_freq, aes(x=rownames(bigram_freq), y = bigram_freq[,1]))
p <- p + geom_bar(fill="#751B1C", stat="identity")
p <- p + ggtitle(paste0("FIORINA bigram sample: \"", word_sample,"\" "))
p <- p + xlab("bigrams")
p <- p + ylab("count")
p <- p + theme(axis.text.x=element_text(face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"))
p <- p + scale_y_continuous(limits=c(0, 10), breaks=c(0, 2, 4, 6, 8, 10))
p <- p + coord_flip()
plot_tax_fiorina <- p
p

```

```{r, echo=FALSE, fig.align="center", fig.height=4, fig.width=6, warning=FALSE, message=FALSE}


#word_sample <- "tax"
a<-bigram_finder(word_sample, sanders_bigrams)
a<-as.data.frame(a)
colnames(a)<-word_sample

bigram_freq<-a

library(ggplot2)

p <- ggplot(bigram_freq, aes(x=rownames(bigram_freq), y = bigram_freq[,1]))
p <- p + geom_bar(fill="#415DA3", stat="identity")
p <- p + ggtitle(paste0("SANDERS bigram sample: \"", word_sample,"\" "))
p <- p + xlab("bigrams")
p <- p + ylab("count")
p <- p + theme(axis.text.x=element_text(face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"))
p <- p + scale_y_continuous(limits=c(0, 10), breaks=c(0,2,4,6,8,10))
p <- p + coord_flip()
plot_tax_sanders <- p

plot_tax_sanders

```

```{r, echo=FALSE, fig.align="center", fig.height=4, fig.width=6, warning=FALSE, message=FALSE}


#word_sample <- "tax"
a<-bigram_finder(word_sample, trump_bigrams)
a<-as.data.frame(a)
colnames(a)<-word_sample

bigram_freq<-a

library(ggplot2)

p <- ggplot(bigram_freq, aes(x=rownames(bigram_freq), y = bigram_freq[,1]))
p <- p + geom_bar(fill="#751B1C", stat="identity")
p <- p + ggtitle(paste0("TRUMP bigram sample: \"", word_sample,"\" "))
p <- p + xlab("bigrams")
p <- p + ylab("count")
p <- p + theme(axis.text.x=element_text(face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"))
p <- p + scale_y_continuous(limits=c(0, 10), breaks=c(0,2,4,6,8,10))
p <- p + coord_flip()
p

```

```{r, echo=FALSE, fig.align="center", fig.height=4, fig.width=6, warning=FALSE, message=FALSE}


#word_sample <- "tax"
a<-bigram_finder(word_sample, clinton_bigrams)
a<-as.data.frame(a)
colnames(a)<-word_sample

bigram_freq<-a

library(ggplot2)

p <- ggplot(bigram_freq, aes(x=rownames(bigram_freq), y = bigram_freq[,1]))
p <- p + geom_bar(fill="#415DA3", stat="identity")
p <- p + ggtitle(paste0("CLINTON bigram sample: \"", word_sample,"\" "))
p <- p + xlab("bigrams")
p <- p + ylab("count")
p <- p + theme(axis.text.x=element_text(face="italic", color="#222222"), axis.text.y=element_text(face="italic", color="#222222"))
p <- p + scale_y_continuous(limits=c(0, 10), breaks=c(0,2,4,6,8,10))
p <- p + coord_flip()
p



```

